From the function-sheaf dictionary to quasicharacters of $p$-adic tori

### 2015-07-21 dictionary.tex
etale mess up, fixed

%%%%%%%%%%%%%%% RINGS AND GROUPS %%%%%%%%%%%%%%%
\newcommand{\FF}{{\mathbb{F}}}
\newcommand{\ZZ}{{\mathbb{Z}}}
\newcommand{\NN}{{\mathbb{N}}}
\newcommand{\CC}{{\mathbb{C}}}
\newcommand{\QQ}{{\mathbb{Q}}}
\newcommand{\RR}{{\mathbb{R}}}
\newcommand{\EE}{\mathbb{\bar Q}_\ell}
\newcommand{\OK}{\mathcal{O}_K}
\newcommand{\OL}{\mathcal{O}_L}
\newcommand{\OO}[1]{\mathcal{O}_{#1}}
\newcommand{\bFq}{\bar{k}}
\newcommand{\Fq}{k}
\newcommand{\Fqm}{k_m}
\newcommand{\EEx}{\EE^\times}
\newcommand{\ZEx}{\mathbb{\bar Z}_\ell^\times}
\newcommand{\Weil}[1]{\mathcal{W}_{#1}}
\newcommand{\m}{{\mathfrak{m}}}
%%%%%%%%%%%%%%% ALGEBRAIC GROUPS %%%%%%%%%%%%%%%
\newcommand{\Gm}[1]{\mathbb{G}_{\hskip-1pt\textbf{m},#1}}
\DeclareMathOperator{\GL}{GL}
\newcommand{\comp}{\Pi} % Component group
%%%%%%%%%%%%%%% NAMED OPERATORS %%%%%%%%%%%%%%%
\DeclareMathOperator{\Gal}{Gal}
\newcommand{\Frob}[1]{\operatorname{F}_{#1}}
\DeclareMathOperator{\Aut}{Aut}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\ord}{ord}
\DeclareMathOperator{\coker}{coker}
\DeclareMathOperator{\Gr}{Gr}
\DeclareMathOperator{\Irrep}{Irrep}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\Ext}{Ext}
\DeclareMathOperator{\Hh}{H}
\DeclareMathOperator{\Res}{Res}
\DeclareMathOperator{\Nm}{Nm}
\DeclareMathOperator{\trace}{Tr}
\DeclareMathOperator{\obj}{obj}
\DeclareMathOperator{\mor}{mor}
\DeclareMathOperator{\Lang}{Lang}
\DeclareMathOperator{\image}{im}
\DeclareMathOperator{\Loc}{Loc}
\DeclareMathOperator{\Tot}{Tot}
\DeclareMathOperator{\Tor}{Tor}
\newcommand{\gal}[1]{{\operatorname{Gal}\hskip-1pt\left( {\bar #1}/#1 \right)}}
\newcommand{\Spec}[1]{{\operatorname{Spec}(#1)}}
%%%%%%%%%%%% MISCELLANEOUS OPERATORS %%%%%%%%%%%%
\newcommand{\sheafHom}{{\mathscr{H}\hskip-4pt{\it o}\hskip-2pt{\it m}}}
\newcommand{\abs}[1]{{\vert #1 \vert}}
\newcommand{\ceq}{{\, :=\, }}
\newcommand{\tq}{{\ \vert\ }}
\newcommand{\iso}{{\ \cong\ }}
\newcommand{\trFrob}[1]{t_{#1}}
\newcommand{\TrFrob}[1]{T_{#1}}
\newcommand{\lTrFrob}[1]{\TrFrob{#1}^2}
%% Limits
\newcommand{\invlim}[1]{\lim\limits_{\overleftarrow{#1}}}
\newcommand{\dirlim}[1]{\lim\limits_{\overrightarrow{#1}}}
\newcommand{\limit}[1]{\mathop{\textsc{lim}}\limits_{#1}}
\newcommand{\colimit}[1]{\mathop{\textsc{colim}}\limits_{#1}}
%% Fonts for quasicharacter sheaves
\newcommand{\cs}[1]{{\mathcal{#1}}}
\newcommand{\gcs}[1]{{\mathcal{\bar #1}}}
\newcommand{\dualgcs}[1]{\gcs{#1}^\dagger}
\newcommand{\dualcs}[1]{\cs{#1}^\dagger}
%% Categories
\newcommand{\CS}{{\mathcal{C\hskip-0.8pt S}}}
\newcommand{\lCS}{{\CS_\text{flb}}}
\newcommand{\lCSiso}[1]{\lCS(#1)_{/\text{iso}}}
\newcommand{\bCS}{{\CS_0}}
\newcommand{\CSiso}[1]{\CS(#1)_{/\text{iso}}}
\newcommand{\bCSiso}[1]{\bCS(#1)_{/\text{iso}}}
\newcommand{\catname}[1]{\normalfont{\textsf{#1}}}
\newcommand{\Sch}[1]{{\catname{Sch}_{/#1}}}
\newcommand{\QCS}{{\mathcal{QC\hskip-0.8pt S}}}
\newcommand{\QCSiso}[1]{\QCS(#1)_{/\text{iso}}}
%% Labeled items
\makeatletter
\newcommand{\labitem}[2]{
\def\@itemlabel{\textbf{#1}}
\item
\def\@currentlabel{#1}\label{#2}}
\makeatother
%% Shorthand for bars
\renewcommand{\bf}{\bar{f}}
\newcommand{\bg}{{\bar{g}}}
\newcommand{\bm}{\bar{m}}
\newcommand{\bG}{\bar{G}}
\newcommand{\bH}{\bar{H}}
\newcommand{\brho}{{\bar\rho}}
\newcommand{\bx}{{\bar{x}}}
%% Spacing control
\newcommand{\tight}[3]{\hspace{-#1pt}{#2}\hspace{-#3pt}}
\newcommand{\GxG}{\text{$G \tight{1}{\times}{1} G$}}
\newcommand{\bGxG}{\text{$\bar{G} \tight{1}{\times}{1} \bar{G}$}}
\newcommand{\bfxf}{\text{$\bar{f} \tight{1}{\times}{1} \bar{f}$}}
\newcommand{\GxxG}{\text{$G \tight{1}{\times}{1} G$}}
\newcommand{\LxL}{\text{$\gcs{L} \tight{0}{\boxtimes}{0} \gcs{L}$}}
\newcommand{\ExE}{\text{$\cs{E}\tight{0}{\boxtimes}{0}\cs{E}$}}
\newcommand{\bExE}{\text{$\gcs{E}\tight{0}{\boxtimes}{0}\gcs{E}$}}
\newcommand{\AxA}{\text{$A \tight{1}{\times}{1} A$}}
\newcommand{\BxB}{\text{$B \tight{1}{\times}{1} B$}}
\newcommand{\GzxGz}{\text{$G^0 \tight{1}{\times}{1} G^0$}}
%% Hyphenation override
\hyphenation{quasi-character}

@@

As Deligne explained in \cite{deligne:SGA4.5}*{Sommes trig.}, if $G$ is a connected commutative algebraic group over a finite field $k$, then the trace of Frobenius provides a bijection between the group $G(\Fq)^*$ of $\ell$-adic characters of $G(\Fq)$ and isomorphism classes of rank-one $\ell$-adic local systems $\mathcal{E}$ on $G$ for which 
\begin{equation}\label{introbox}
m^* \cs{E} \iso \cs{E} \boxtimes \cs{E},
\end{equation}
where $m : G\times G\to G$ is the multiplication map.
%
If one wishes to make a category from this class of local systems, one is led to consider morphisms $\cs{E} \to \cs{E}'$ of sheaves which are compatible with particular choices of \eqref{introbox} for $\cs{E}$ and $\cs{E'}$. 
{\it A priori}, the composition $\cs{E} \to \cs{E}' \to \cs{E}''$ of two such morphisms need not be compatible with the choices of \eqref{introbox} for $\cs{E}$ and $\cs{E}''$.
 However, for connected $G$ the isomorphism \eqref{introbox} is unique, if it exists, and there is no impediment to making the dictionary categorical.

If $G$ is a commutative algebraic group over $\Fq$ which is not connected, however, then the isomorphism \eqref{introbox} need not be unique. In order to track the choice of isomorphism, consider the category $\bCS(G)$ of pairs $(\cs{E},\mu_\cs{E})$ where $\cs{E}$ is a rank-one local system on $G$ and $\mu_\cs{E} : m^*\cs{E} \to \cs{E}\boxtimes\cs{E}$ is a chosen isomorphism of local systems on $G\times G$. 
In this case, the trace of Frobenius provides an epimorphism from isomorphism classes of objects in $\bCS(G)$ to characters of $G(\Fq)$, but the epimomorphism need not be injective; consequently,
every character of $G(\Fq)$ may be geometrized as a pair $(\cs{E},\mu_\cs{E})$ but perhaps not uniquely.
Indeed, it follows from a special case of the main result of this paper that the kernel of the trace of Frobenius $T_{G} : \bCS(G) \to G(\Fq)^*$ trivial if and only if the \'etale group scheme of connected components of $G$ is finite cyclic.
Nevertheless, the defect in the dictionary for characters of commutative algebraic groups over finite fields may be addressed with the following observation: if $(\cs{E},\mu_\cs{E})$ and $(\cs{E}',\mu_\cs{E}')$ determine the same character of $G(\Fq)$ then $\cs{E}\iso \cs{E}'$ as local systems on $G$.

In order to extend the dictionary from commutative algebraic groups over finite fields to smooth commutative group schemes $G$ over $\Fq$,
we replace the local system $\cs{E}$ on $G$ with a Weil local system while retaining the extra structure $\mu_\cs{E}$. 
In this way we are led to the category $\CS(G)$ of {\it character sheaves} on $G$ (\S\ref{ssec:category}):
objects in $\CS(G)$ are triples $(\gcs{L}, \mu,\phi)$, where $\gcs{L}$ is a rank-one local system on $\bG := G\times_{\Spec{\Fq}} \Spec{\bFq}$
and $\phi : \Frob{G}^* \gcs{L}\to \gcs{L}$ and $\mu : \bm^* \gcs{L} \iso \gcs{L} \boxtimes \gcs{L}$ are isomorphisms of sheaves satisfying certain compatibility conditions;
morphisms in $\CS(G)$ are then morphisms of Weil sheaves which are compatible with the extra structure.
This paper establishes the basic properties of $\CS(G)$ by finding the relation between character sheaves on $G$ and characters of $G(\Fq)$.

We begin our study of $\CS(G)$ by returning to the case when $G$ is a connected commutative algebraic group over $\Fq$, revisiting Deligne's function-sheaf dictionary (\S\ref{sec:category}).
We consider character sheaves that arise via base change to $\bFq$ from local systems on $G$ (\S\ref{ssec:descentG}) and
those that appear in a pushforward from a constant sheaf along a discrete isogeny $H \to G$ (\S\ref{ssec:discrete_isogenies}).  
While these constructions make sense even for non-connected $G$, in the connected case we show that every character sheaf can be described in both of these ways (\S\ref{ssec:connected}).  
We use this fact to prove that $\TrFrob{G}$ is an isomorphism for connected commutative algebraic groups $G$
and to determine the automorphism groups of character sheaves on such $G$. 
%These results are all well known.

Next, we consider character sheaves on \'etale commutative group schemes $G$ over $\Fq$ (\S\ref{sec:etale}).
\'Etale group schemes form a counterpoint to connected algebraic groups, since the component group of any smooth group scheme is an \'etale group scheme.
%etale group schemes are algebraic groups only when they are finite.
Our key tools for understanding the trace of Frobenius in the \'etale case 
are a reinterpretation of $\CS(G)$ in terms of stalks (\S\ref{ssec:stalks}) and 
%a relationship with 
 the Hochschild-Serre spectral sequence (\S\ref{ssec:E}) for $\Weil{} \ltimes \bG$, where $\Weil{} \subset \Gal(\Fq/\Fq)$ is the Weil group for $\Fq$.
We define (\S\ref{ssec:S}) an isomorphism $S_G$ from $\CSiso{G}$ to the second cohomology of the total space of the spectral sequence
\[
E_2^{p,q} \ceq \Hh^p(\Weil{}, \Hh^q(\bG, \EEx)) \Rightarrow \Hh^{p+q}(\Weil{} \ltimes \bG, \EEx).
\]
Paired with the short exact sequence
\[
  0 \to \Hh^0(\Weil{},\Hh^2(\bG,\EEx)) \to \Hh^2(E^\bullet_G) \to \Hh^1(\Weil{},\Hh^1(\bG,\EEx)) \to 0
\]
arising from the spectral sequence, the isomorphism $S_G$ allows us to show (\S\ref{ssec:SandT}) that 
the group homomorphism 
\[
\TrFrob{G}: \CSiso{G} \to G(\Fq)^*
\]
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$.
The necessity of using Weil local systems on $G$ in the definition of $\CS(G)$ already appears here:  if one were to use local systems on $G$ instead, the group homomorphism $\TrFrob{G}$ would not then be surjective (\S \ref{ssec:bS}). 
One can show, using the K\"unneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is cyclic, so the kernel of $\TrFrob{G}$ is non-trivial in general.
%The fact that $\TrFrob{G}$ may not be an isomorphism should not be too troubling: if quasicharacter sheaves $(\gcs{L}, \mu,\phi)$ and $(\gcs{L}', \mu',\phi')$ determine the same character of $G(\Fq)$ then $(\gcs{L},\phi)$ and $(\gcs{L}',\phi')$ are isomorphic as Weil sheaves.

%The appearance of $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ in these two cases is interesting: on the one hand, when $G$ is a connected commutative algebraic group over $\Fq$, isomorphism classes of quasicharacter sheaves on $G$ are measured by $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ and the trace of Frobenius gives an isomorphism $\TrFrob{G}: \CSiso{G} \to \Hom(G(\Fq),\EEx)$; on the other hand, if $G$ an etale commutative group scheme over $\Fq$, then $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ measures the kernel of $\TrFrob{G}: \CSiso{G} \to \Hom(G(\Fq),\EEx)$. 
%This observation makes plain the fact that  . . .
%Indeed, quasicharacter sheaves on \'etale commutative group schemes that are trivialized by discrete isogenies are invisible to the trace of Frobenius.

Having understood $\CS(G)$ in two extreme cases --
for connected commutative algebraic groups and for \'etale commutative group schemes -- we return to the case of smooth commutative group schemes (\S\ref{sec:main}) using the component group sequence
\[
0 \to G^0 \to G \to \pi_0(G) \to 0.
\]
Using pullbacks of character sheaves we obtain a diagram
\[
  \begin{tikzcd}[row sep=20, column sep=20]
    0 \rar & \CSiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \CSiso{G} \rar \dar{\TrFrob{G}} & \CSiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar & G(\Fq)^* \rar & G^0(\Fq)^* \rar & 0.
  \end{tikzcd}
\]
We show that the rows of this diagram are exact (\S\S\ref{ssec:restriction},\ref{ssec:component}), so we may apply the snake lemma to prove the main theorem of the paper,
\begin{theorem*}[{Thm. \ref{thm:snake}}]
If $G$ is a smooth commutative group scheme over $\Fq$ then the trace of Frobenius gives a short exact sequence
\[
\begin{tikzcd}
0 \arrow{r} & \Hh^2(\pi_0(\bG),\EEx)^{\Weil{}} \arrow{r} & \CSiso{G} \arrow{r}{\TrFrob{G}} & G(\Fq)^* \arrow{r} & 0.
\end{tikzcd}
\]
\end{theorem*}
\noindent
Thus, unless the component group scheme $\pi_0(\bG)$ is cyclic, category $\CS(G)$ contains non-trivial character sheaves lying in the kernel of the trace of Frobenius and therefore lost in translation to characters of $G(\Fq)$.
%These {\it invisible character sheaves} are parametrized by $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}$.

@@

\medskip
\noindent\textbf{Application to quasicharacters of $p$-adic tori and abelian varieties.\ }
As indicated above, our interest in the function-sheaf dictionary for smooth commutative group schemes
over finite fields comes from an application to $p$-adic representation theory,
specifically to quasicharacters (\S\ref{ssec:quasicharacters}) of $p$-adic tori.
However, we found that our method of passing from $p$-adic tori to group schemes over $\Fq$ applies more generally to
any local field $K$ with finite residue field $\Fq$ and any commutative algebraic group $X_K$ that admits a N\'eron model $X$.
This class of groups includes abelian varieties and unipotent $K$-wound groups in addition to the tori we initially considered.

In this paper we show that quasicharacters of $X(K)$ are geometrized and categorified by quasicharacter sheaves on the
Greenberg transform $\Gr_R(X)$ of the N\'eron model $X$.
Although not locally of finite type, $\Gr_R(X)$ is a commutative group scheme over $\Fq$ and also a projective limit of smooth commutative group schemes $\Gr^R_n(X)$.
This structure allows us to adapt our work on character sheaves on smooth group schemes over finite fields to construct (\S\ref{ssec:CS_on_GN}) a category $\QCS(X)$
of quasicharacter sheaves for $X$, which are certain sheaves on $\Gr_R(X)\times_{\Spec{\Fq}} \Spec{\bFq}$, with extra structure.
The ability to generalize the function-sheaf dictionary to non-connected group schemes plays a crucial role in this application.

>>>>

%%%%%%%%%%%%%%% RINGS AND GROUPS %%%%%%%%%%%%%%%
\newcommand{\FF}{{\mathbb{F}}}
\newcommand{\ZZ}{{\mathbb{Z}}}
\newcommand{\NN}{{\mathbb{N}}}
\newcommand{\CC}{{\mathbb{C}}}
\newcommand{\QQ}{{\mathbb{Q}}}
\newcommand{\RR}{{\mathbb{R}}}
\newcommand{\EE}{\mathbb{\bar Q}_\ell}
\newcommand{\OK}{\mathcal{O}_K}
\newcommand{\OL}{\mathcal{O}_L}
\newcommand{\OO}[1]{\mathcal{O}_{#1}}
\newcommand{\bFq}{\bar{k}}
\newcommand{\Fq}{k}
\newcommand{\Fqm}{k_m}
\newcommand{\EEx}{\EE^\times}
\newcommand{\ZEx}{\mathbb{\bar Z}_\ell^\times}
\newcommand{\Weil}[1]{\mathcal{W}_{#1}}
\newcommand{\m}{{\mathfrak{m}}}
%%%%%%%%%%%%%%% ALGEBRAIC GROUPS %%%%%%%%%%%%%%%
\newcommand{\Gm}[1]{\mathbb{G}_{\hskip-1pt\textbf{m},#1}}
\DeclareMathOperator{\GL}{GL}
\newcommand{\comp}{\Pi} % Component group
%%%%%%%%%%%%%%% NAMED OPERATORS %%%%%%%%%%%%%%%
\DeclareMathOperator{\Gal}{Gal}
\newcommand{\Frob}[1]{\operatorname{Fr}_{#1}}
\DeclareMathOperator{\Aut}{Aut}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\ord}{ord}
\DeclareMathOperator{\coker}{coker}
\DeclareMathOperator{\Gr}{Gr}
\DeclareMathOperator{\Irrep}{Irrep}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\Ext}{Ext}
\DeclareMathOperator{\Hh}{H}
\DeclareMathOperator{\Res}{Res}
\DeclareMathOperator{\Nm}{Nm}
\DeclareMathOperator{\trace}{Tr}
\DeclareMathOperator{\obj}{obj}
\DeclareMathOperator{\mor}{mor}
\DeclareMathOperator{\Lang}{Lang}
\DeclareMathOperator{\image}{im}
\DeclareMathOperator{\Loc}{Loc}
\DeclareMathOperator{\Tot}{Tot}
\DeclareMathOperator{\Tor}{Tor}
\newcommand{\gal}[1]{{\operatorname{Gal}\hskip-1pt\left( {\bar #1}/#1 \right)}}
\newcommand{\Spec}[1]{{\operatorname{Spec}(#1)}}
%%%%%%%%%%%% MISCELLANEOUS OPERATORS %%%%%%%%%%%%
\newcommand{\sheafHom}{{\mathscr{H}\hskip-4pt{\it o}\hskip-2pt{\it m}}}
\newcommand{\abs}[1]{{\vert #1 \vert}}
\newcommand{\ceq}{{\, :=\, }}
\newcommand{\tq}{{\ \vert\ }}
\newcommand{\iso}{{\ \cong\ }}
\newcommand{\trFrob}[1]{t_{#1}}
\newcommand{\TrFrob}[1]{\operatorname{Tr}_{#1}}
\newcommand{\lTrFrob}[1]{\TrFrob{#1}^2}
%% Limits
\newcommand{\invlim}[1]{\lim\limits_{\overleftarrow{#1}}}
\newcommand{\dirlim}[1]{\lim\limits_{\overrightarrow{#1}}}
\newcommand{\limit}[1]{\mathop{\textsc{lim}}\limits_{#1}}
\newcommand{\colimit}[1]{\mathop{\textsc{colim}}\limits_{#1}}
%% Fonts for quasicharacter sheaves
\newcommand{\cs}[1]{{\mathcal{#1}}}
\newcommand{\gcs}[1]{{\mathcal{\bar #1}}}
\newcommand{\dualgcs}[1]{\gcs{#1}^\dagger}
\newcommand{\dualcs}[1]{\cs{#1}^\dagger}
%% Categories
\newcommand{\CS}{{\mathcal{C\hskip-0.8pt S}}}
\newcommand{\lCS}{{\CS_\text{flb}}}
\newcommand{\lCSiso}[1]{\lCS(#1)_{/\text{iso}}}
\newcommand{\bCS}{{\CS_0}}
\newcommand{\CSiso}[1]{\CS(#1)_{/\text{iso}}}
\newcommand{\bCSiso}[1]{\bCS(#1)_{/\text{iso}}}
\newcommand{\catname}[1]{\normalfont{\textsf{#1}}}
\newcommand{\Sch}[1]{{\catname{Sch}_{/#1}}}
\newcommand{\QCS}{{\mathcal{QC\hskip-0.8pt S}}}
\newcommand{\QCSiso}[1]{\QCS(#1)_{/\text{iso}}}
%% Labeled items
\makeatletter
\newcommand{\labitem}[2]{
\def\@itemlabel{\textbf{#1}}
\item
\def\@currentlabel{#1}\label{#2}}
\makeatother
%% Shorthand for bars
\renewcommand{\bf}{\bar{f}}
\newcommand{\bg}{{\bar{g}}}
\newcommand{\bm}{\bar{m}}
\newcommand{\bG}{\bar{G}}
\newcommand{\bH}{\bar{H}}
\newcommand{\brho}{{\bar\rho}}
\newcommand{\bx}{{\bar{x}}}
%% Spacing control
\newcommand{\tight}[3]{\hspace{-#1pt}{#2}\hspace{-#3pt}}
\newcommand{\GxG}{\text{$G \tight{1}{\times}{1} G$}}
\newcommand{\bGxG}{\text{$\bar{G} \tight{1}{\times}{1} \bar{G}$}}
\newcommand{\bfxf}{\text{$\bar{f} \tight{1}{\times}{1} \bar{f}$}}
\newcommand{\GxxG}{\text{$G \tight{1}{\times}{1} G$}}
\newcommand{\LxL}{\text{$\gcs{L} \tight{0}{\boxtimes}{0} \gcs{L}$}}
\newcommand{\ExE}{\text{$\cs{E}\tight{0}{\boxtimes}{0}\cs{E}$}}
\newcommand{\bExE}{\text{$\gcs{E}\tight{0}{\boxtimes}{0}\gcs{E}$}}
\newcommand{\AxA}{\text{$A \tight{1}{\times}{1} A$}}
\newcommand{\BxB}{\text{$B \tight{1}{\times}{1} B$}}
\newcommand{\GzxGz}{\text{$G^0 \tight{1}{\times}{1} G^0$}}
%% Hyphenation override
\hyphenation{quasi-character}

@@

As Deligne explained in \cite{deligne:SGA4.5}*{Sommes trig.}, if $G$ is a connected commutative algebraic group over a finite field $k$, then the trace of Frobenius provides a bijection between the group $G(\Fq)^*$ of $\ell$-adic characters of $G(\Fq)$ and isomorphism classes of those rank-one $\ell$-adic local systems $\mathcal{E}$ on $G$ for which 
\begin{equation}\label{introbox}
m^* \cs{E} \iso \cs{E} \boxtimes \cs{E},
\end{equation}
where $m : G\times G\to G$ is the multiplication map.
%
If one wishes to make a category from this class of local systems, one is led to consider morphisms $\cs{E} \to \cs{E}'$ of sheaves which are compatible with particular choices of \eqref{introbox} for $\cs{E}$ and $\cs{E'}$. 
{\it A priori}, the composition $\cs{E} \to \cs{E}' \to \cs{E}''$ of two such morphisms need not be compatible with the choices of \eqref{introbox} for $\cs{E}$ and $\cs{E}''$.
 However, for connected $G$ the isomorphism \eqref{introbox} is unique, if it exists, and there is no impediment to making the dictionary categorical.

If $G$ is a commutative algebraic group over $\Fq$ which is not connected, however, then the isomorphism \eqref{introbox} need not be unique. In order to track the choice of isomorphism, consider the category $\bCS(G)$ of pairs $(\cs{E},\mu_\cs{E})$ where $\cs{E}$ is a rank-one local system on $G$ and $\mu_\cs{E} : m^*\cs{E} \to \cs{E}\boxtimes\cs{E}$ is a chosen isomorphism of local systems on $G\times G$. 
In this case, the trace of Frobenius provides an epimorphism from isomorphism classes of objects in $\bCS(G)$ to characters of $G(\Fq)$, but the epimomorphism need not be injective; consequently,
every character of $G(\Fq)$ may be geometrized as a pair $(\cs{E},\mu_\cs{E})$ but perhaps not uniquely.
Indeed, it follows from a special case of the main result of this paper that the kernel of the trace of Frobenius $\bCS(G) \to G(\Fq)^*$ trivial if and only if the  group scheme of connected components of $G$ is cyclic.
The defect in the function-sheaf dictionary for characters of commutative algebraic groups over finite fields may be addressed with the following observation: if $(\cs{E},\mu_\cs{E})$ and $(\cs{E}',\mu_\cs{E}')$ determine the same character of $G(\Fq)$ then $\cs{E}\iso \cs{E}'$ as local systems on $G$.

Motivated by an application to quasicharacters of algebraic tori over local fields, in this paper we extend the function-sheaf dictionary from commutative algebraic groups over finite fields to smooth commutative group schemes $G$ over $\Fq$.
In order to do this, we replace the local system $\cs{E}$ on $G$ with a Weil local system while retaining the extra structure $\mu_\cs{E}$. 
In this way we are led to the category $\CS(G)$ of {\it character sheaves} on $G$ (\S\ref{ssec:category}):
objects in $\CS(G)$ are triples $(\gcs{L}, \mu,\phi)$, where $\gcs{L}$ is a rank-one local system on $\bG := G\times_{\Spec{\Fq}} \Spec{\bFq}$
and $\phi : \Frob{G}^* \gcs{L}\to \gcs{L}$ and $\mu : \bm^* \gcs{L} \iso \gcs{L} \boxtimes \gcs{L}$ are isomorphisms of sheaves satisfying certain compatibility conditions;
morphisms in $\CS(G)$ are then morphisms of Weil sheaves which are compatible with the extra structure.
This paper establishes the basic properties of category $\CS(G)$, using the group homomorphism 
\[
\TrFrob{G}: \CSiso{G} \to G(\Fq)^*
\]
provided by the trace of Frobenius  to find the relation between character sheaves on $G$ and characters of $G(\Fq)$.
Then we return to our motivating application and use character sheaves to geometrize and categorify quasicharacters algebraic tori over local fields.
% (\S \ref{sec:applications}).

We begin our study of category $\CS(G)$ by returning to the case when $G$ is a connected commutative algebraic group over $\Fq$, revisiting Deligne's function-sheaf dictionary (\S\ref{sec:category}).
We consider character sheaves that arise via base change to $\bFq$ from local systems on $G$ (\S\ref{ssec:descentG}) and
those that appear in a pushforward from a constant sheaf along a discrete isogeny $H \to G$ (\S\ref{ssec:discrete_isogenies}).  
While these constructions make sense even for non-connected $G$, in the connected case we show that every character sheaf can be described in both of these ways (\S\ref{ssec:connected}).  
We use this fact to prove that $\TrFrob{G} : \CS(G) \to G(\Fq)^*$ is an isomorphism for connected commutative algebraic groups $G$.
We also determine the automorphism groups of character sheaves on such $G$. 
These facts are well known.

Next, we consider character sheaves on \'etale commutative group schemes $G$ over $\Fq$ (\S\ref{sec:etale}).
\'Etale group schemes form a counterpoint to connected algebraic groups, since the component group of any smooth group scheme is an \'etale group scheme.
%etale group schemes are algebraic groups only when they are finite.
Our key tools for understanding the trace of Frobenius in the \'etale case 
are a reinterpretation of $\CS(G)$ in terms of stalks (\S\ref{ssec:stalks}) and 
%a relationship with 
 the Hochschild-Serre spectral sequence (\S\ref{ssec:E}) for $\Weil{} \ltimes \bG$, where $\Weil{} \subset \Gal(\Fq/\Fq)$ is the Weil group for $\Fq$.
We define (\S\ref{ssec:S}) an isomorphism $S_G$ from $\CSiso{G}$ to the second cohomology of the total space of the spectral sequence
\[
E_2^{p,q} \ceq \Hh^p(\Weil{}, \Hh^q(\bG, \EEx)) \Rightarrow \Hh^{p+q}(\Weil{} \ltimes \bG, \EEx).
\]
Paired with the short exact sequence
\[
  0 \to \Hh^0(\Weil{},\Hh^2(\bG,\EEx)) \to \Hh^2(E^\bullet_G) \to \Hh^1(\Weil{},\Hh^1(\bG,\EEx)) \to 0
\]
arising from the spectral sequence, the isomorphism $S_G$ allows us to show (\S\ref{ssec:SandT}) that the group homomorphism $\TrFrob{G}: \CSiso{G} \to G(\Fq)^*$ is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$.
The necessity of using Weil local systems on $G$ in the definition of $\CS(G)$ already appears here:  if one were to use local systems on $G$ instead, the group homomorphism $\TrFrob{G}$ would not then be surjective (\S \ref{ssec:bS}). 
Moreover, as examples show (\S \ref{eg:}), the kernel of $\TrFrob{G}$ is non-trivial in general.
%The fact that $\TrFrob{G}$ may not be an isomorphism should not be too troubling: if quasicharacter sheaves $(\gcs{L}, \mu,\phi)$ and $(\gcs{L}', \mu',\phi')$ determine the same character of $G(\Fq)$ then $(\gcs{L},\phi)$ and $(\gcs{L}',\phi')$ are isomorphic as Weil sheaves.

%The appearance of $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ in these two cases is interesting: on the one hand, when $G$ is a connected commutative algebraic group over $\Fq$, isomorphism classes of quasicharacter sheaves on $G$ are measured by $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ and the trace of Frobenius gives an isomorphism $\TrFrob{G}: \CSiso{G} \to \Hom(G(\Fq),\EEx)$; on the other hand, if $G$ an \'etale commutative group scheme over $\Fq$, then $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ measures the kernel of $\TrFrob{G}: \CSiso{G} \to \Hom(G(\Fq),\EEx)$. 
%This observation makes plain the fact that  . . .
%Indeed, quasicharacter sheaves on \'etale commutative group schemes that are trivialized by discrete isogenies are invisible to the trace of Frobenius.

Having understood $\CS(G)$ in two extreme cases --
for connected commutative algebraic groups and for \'etale commutative group schemes -- we turn to the case of smooth commutative group schemes (\S\ref{sec:main}) using the component group sequence
\[
0 \to G^0 \to G \to \pi_0(G) \to 0.
\]
Using pullbacks of character sheaves we obtain a diagram
\[
  \begin{tikzcd}[row sep=20, column sep=20]
    0 \rar & \CSiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \CSiso{G} \rar \dar{\TrFrob{G}} & \CSiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar & G(\Fq)^* \rar & G^0(\Fq)^* \rar & 0.
  \end{tikzcd}
\]
We show that the rows of this diagram are exact (\S\S\ref{ssec:restriction},\ref{ssec:component}), so we may apply the snake lemma to prove the main theorem of the paper,
\begin{theorem*}[{Thm. \ref{thm:snake}}]
If $G$ is a smooth commutative group scheme over $\Fq$ then the trace of Frobenius gives a short exact sequence
\[
\begin{tikzcd}
0 \arrow{r} & \Hh^2(\pi_0(\bG),\EEx)^{\Weil{}} \arrow{r} & \CSiso{G} \arrow{r}{\TrFrob{G}} & G(\Fq)^* \arrow{r} & 0.
\end{tikzcd}
\]
\end{theorem*}
\noindent
Thus, unless the component group scheme $\pi_0(\bG)$ is cyclic, category $\CS(G)$ contains non-trivial character sheaves lying in the kernel of the trace of Frobenius and therefore lost in translation to characters of $G(\Fq)$.
%These {\it invisible character sheaves} are parametrized by $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}$.

@@

\medskip
\noindent\textbf{Application to quasicharacters of $p$-adic tori and abelian varieties.\ }
As indicated above, our interest in the function-sheaf dictionary for smooth commutative group schemes
over finite fields comes from an application to $p$-adic representation theory,
specifically to quasicharacters (\S\ref{ssec:quasicharacters}) of $p$-adic tori.
However, we found that our method of passing from $p$-adic tori to group schemes over $\Fq$ applies more generally to
any local field $K$ with finite residue field $\Fq$ and to any commutative algebraic group over $K$ that admits a N\'eron model $X$.
This class of algebraic groups over $K$ includes abelian varieties and unipotent $K$-wound groups, in addition to the algebraic tori we initially considered.

In this paper we show that if $X$ is as above then quasicharacters of $X(K)$ are geometrized and categorified by character sheaves on the
Greenberg transform $\Gr_R(X)$ of the N\'eron model $X$.
Although not locally of finite type, $\Gr_R(X)$ is a commutative group scheme over $\Fq$ and also a projective limit of smooth commutative group schemes $\Gr^R_n(X)$.
This structure allows us to adapt our work on character sheaves on smooth group schemes over finite fields to construct (\S\ref{ssec:CS_on_GN}) a category $\QCS(X)$
of {\it quasicharacter sheaves} for $X$, which are certain sheaves on $\Gr_R(X)\times_{\Spec{\Fq}} \Spec{\bFq}$, with extra structure.
The ability to generalize the function-sheaf dictionary to non-connected group schemes plays a crucial role in this application.
